---
title: "Using `RTCGA` package to download clinical data that are included in `RTCGA.clinical` package"
subtitle: "Date of datasets release: 2015-08-21"
author: "Marcin Kosinski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RTCGA to download clinical data as included in RTCGA.clinical}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(comment="", message=FALSE, warning = FALSE, tidy.opts=list(keep.blank.line=TRUE, width.cutoff=150), options(width=150), eval = FALSE)
```

# RTCGA package

> The Cancer Genome Atlas (TCGA) Data Portal provides a platform for researchers to search, download, and analyze data sets generated by TCGA. It contains clinical information, genomic characterization data, and high level sequence analysis of the tumor genomes. The key is to understand genomics to improve cancer care.

`RTCGA` package offers download and integration of the variety and volume of TCGA data using patient barcode key, what enables easier data possession. This may have a benefcial infuence on  development of science and improvement of patients' treatment. `RTCGA` is an open-source R package, available to download from Bioconductor 

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("RTCGA")
```

or from github
```{r, eval=FALSE}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
biocLite("RTCGA/RTCGA")
```

Furthermore, `RTCGA` package transforms TCGA data into form which is convenient to use in R statistical package. Those data transformations can be a part of statistical analysis pipeline which can be more reproducible with `RTCGA`.

Use cases and examples are shown in `RTCGA` packages vignettes:
```{r, eval=FALSE}
browseVignettes("RTCGA")
```

# How to download clinical data to gain the same datasets as in RTCGA.clinical package?

There are many available date times of TCGA data releases. To see them all just type:
```{r, eval=FALSE}
library(RTCGA)
checkTCGA('Dates')
```

Version 20150821.1.0 of `RTCGA.clinical` package contains clinical datasets which were released `2015-08-21`. They were downloaded in the following way (which is mainly copied from [http://marcinkosinski.github.io/RTCGA/](http://marcinkosinski.github.io/RTCGA/):

## Available cohorts

All cohort names can be checked using:
```{r, eval=FALSE}
(cohorts <- infoTCGA() %>% 
   rownames() %>% 
   sub("-counts", "", x=.))
```

For all cohorts the following code downloads the RPPA data.

## Downloading clinical files
```{r, eval=FALSE}
# dir.create( "data2" ) # name of a directory in which data will be stored
releaseDate <- "2015-08-21"
sapply( cohorts, function(element){
tryCatch({
downloadTCGA( cancerTypes = element, 
              destDir = "data2", 
              date = releaseDate )},
error = function(cond){
   cat("Error: Maybe there weren't mutations data for ", element, " cancer.\n")
}
)
})
```

## Reading downloaded clinical dataset

### Shortening paths and directories 

```{r, eval=FALSE}
list.files( "data2") %>% 
   file.path( "data2", .) %>%
   file.rename( to = substr(.,start=1,stop=50))
```


### Removing `NA` files from data2 folder

If there were not clinical data for some cohorts we should remove corresponding `NA` files.

```{r, eval=FALSE}
list.files( "data2") %>%
   file.path( "data2", .) %>%
   sapply(function(x){
      if (x == "data2/NA")
         file.remove(x)      
   })
```

### Paths to clinical data

Below is the code that removes unneeded "MANIFEST.txt" file from each clinical cohort folder.

```{r}
list.files( "data2") %>% 
   file.path( "data2", .) %>%
   sapply(function(x){
      file.path(x, list.files(x)) %>%
         grep(pattern = "MANIFEST.txt", x = ., value=TRUE) %>%
         file.remove()
      })
```

Below is the code that automatically assigns paths to files for all clinical files for all available cohorts types downloaded to `data2` folder.

```{r}
cohorts %>%
	sapply(function(z){
		list.files("data2") %>%
			file.path("data2", .) %>%
			grep(paste0("_",z,"\\."), x = ., value = TRUE) %>%
			file.path(., list.files(.)) %>%
			grep("clin.merged.txt", x = ., value = TRUE) %>%
			assign(value = .,
						 x = paste0(z, ".clin.path"),
						 envir = .GlobalEnv)
	})
```

### Reading clinical data using `readTCGA`

Because of the fact that clinical data are transposed in downloaded files, there has been prepared special function `readTCGA` to read and transpose data automatically. Code is below

```{r, eval=FALSE}
ls() %>%
   grep("clin\\.path", x = ., value = TRUE) %>% 
   sapply(function(element){
      tryCatch({
         readTCGA(get(element, envir = .GlobalEnv),
               dataType = "clinical") %>%
         assign(value = .,
                x = sub("\\.path", "", x = element),
                envir = .GlobalEnv )
      }, error = function(cond){
         cat(element)
      }) 
     invisible(NULL)
    }    
)
```

# Saving clinical data to `RTCGA.clinical` package


```{r, eval=FALSE}
grep( "clin", ls(), value = TRUE) %>%
   grep("path", x=., value = TRUE, invert = TRUE) %>%
   cat( sep="," ) #can one to it better? as from use_data documentation:
   # ...	Unquoted names of existing objects to save
   devtools::use_data(ACC.clin,BLCA.clin,BRCA.clin,CESC.clin,
   									 CHOL.clin,COAD.clin,COADREAD.clin,DLBC.clin,
   									 ESCA.clin,FPPP.clin,GBM.clin,GBMLGG.clin,
   									 HNSC.clin,KICH.clin,KIPAN.clin,KIRC.clin,
   									 KIRP.clin,LAML.clin,LGG.clin,LIHC.clin,
   									 LUAD.clin,LUSC.clin,MESO.clin,OV.clin,
   									 PAAD.clin,PCPG.clin,PRAD.clin,READ.clin,
   									 SARC.clin,SKCM.clin,STAD.clin,STES.clin,
   									 TGCT.clin,THCA.clin,THYM.clin,UCEC.clin,
   									 UCS.clin,UVM.clin,
                      compress="xz")
```
